A First Effort for a Distributed Segment-based Approach on Topology-agnostic Self-Assembled Nano Networks

### 2013-09-18 trunk/PAPER/introduction.tex
lot of fixes

Exploring long term alternatives to the CMOS technology is gaining
more and more revelance as the scaling trend of such devices keeps
introducing new challenges. Power density, defect tolerance, testing
costs and wire delays are only a few of the many critical aspects
involved~\cite{itrs12}. While software parallelism and multicore
approaches~\cite{horowitz2004, mudge2001, powell2009} are partially mitigating the impact of such
constraints on performances, it is likely that the growing computing demand will
eventually need even more radical architectural modifications and new
paradigms in order to address the Computer Design challenges of the next
decades.

In the last years, Self-assembled nanoscale architectures~\cite{winfree1998, yan2003}
are emerging as a promising technology due their tera/peta scale of
integration, defect tolerance tolerance and huge potential computing
capabilities. These technologies are certainly still at
their early stage of development, however different laboratory demos and
prof-of-concepts architectures have been presented~\cite{patwardhan2004, patwardhan2006, patwardhan2006_1, pistol2009}.
The main idea behind this approach is to exploit the physical regularity and
stabiliy of DNA structures in order to create a scaffold onto which
nano-devices (e.g. nanowires and nanotubes~\cite{bachtold2001, tans1998, cui2001}) can be
attached. This can be achieved by designing approapriate complementary DNA tags for
each terminal to be placed, so that a nano device will be attached
only where its own DNA tag matches a complementary tag on the DNA grid
scaffold (see figure~\ref{fig:dna_tag}.
\begin{figure}
  \centering
    \includegraphics[width=0.40\textwidth]{pictures/dna2b.eps}
  \caption{DNA tag matching}
  \label{fig:dna_tag}
\end{figure}
A detailed description of the chemical properties involved is far
beyond the scope of this paper (see also~\cite{braun1998, seeman1999}), so we will focus on
the three main challenges that this new fabrication process introduces
in Computer Design: (i)limited node complexity, (ii) large scale
randomness and (iii) high defect rates.

The \emph{limited node complexity} aspect is directly
related to the use of complementary DNA tags in order to place circuit
components. In a traditional CMOS process the complexity is introduced
using a photolithography mask, so that larger (more complex) circuits
only require larger masks; conversely, self-assembly achieves
complexity by increasing the number of unique DNA tags, since more
different tags means having more control on component placement. Ideally, by specifying a single
and unique tag for each nano transistor terminal, we could exactly
choose where each component would be placed. But the number of DNA
symbols forming the DNA sequence is limited (sequence of 4 nucleobases
G,A,T,C) and so creating many different tags (of a given lenght) would
mean make them more similar to each other, increasing the probability
of incorrect/partial matching. To avoid this problem we should limit
the number of unique tags, thus limiting the complexity at each node.

\emph{Large scale randomness} and \emph{high defect rates} are the other two fundaments
conditions of self-assembled technology: if, as seen, some complexity
can be introduced by designing tag sequences in the DNA grid of each
node, we have no control of where each of these grids will be placed
on the whole network. Further, as a consequence, we cannot guarantee
other typical properties of regular networks, like nodes being
connected to 4 neighbors, having a determined orientation and so on.

These aspects of a DNA-self assembly process lead to some important
architectural implications to be addressed when approaching to the
Computer Design: computational model must be founded on a distributed
network of small computing and storage nodes, randomly placed and
interconnected (see figure~\ref{fig:nana}).  Proof-of-concept of such kind of computational model
can be found in~\cite{patwardhan2006_1}, where instructions and data operands travel
on packets and routed using a topology-agnostic strategy that avoids
deadlock, since no particular regularity can be assumed in such
network.

\begin{figure}
  \centering
    \includegraphics[width=0.40\textwidth]{pictures/dna1.eps}
  \caption{Irregular architecture of computing and storage elements}
  \label{fig:nana}
\end{figure}

In this paper we introduce DiSR, a first effort for a Distributed
Segment-based Routing in large scale DNA self assembled networks.
Given the irregular/high-defect rates scenario assumed as background, is essential
avoiding non-scalable or resource hungry solutions such as
centralized off-chip algorithms, tree-based approaches, virtual channels
or expensive implementations based on hardware redundancy.
Altought we will assume an architecture similar to the one depicted
in~\ref{fig:nana} as background, it
is important to underline that the proposed approach to topology
discovery is more general and not stricly dependent on any of these
architectures.

@@

The contribution presented in this paper is to bring the advantages of
segment-based turn prohibition without requiring any topology graph,
external defect map or centralized algorithm execution.  While
maintaining similar properties to segment-based routing, the DiSR
approach is not intended to discover the "optimal" segment choice
(ideally reachable with the knowledge of the topology graph) but just
to demonstrate a concrete model that can fit into such complex,
irregular and large sized topologies.

The paper is organized as follows: in Section~\ref{sec:disr_concepts}
we introduce the main elements of the DiSR approach, describing  
a draft of the DiSR model in Section~\ref{sec:disr_model}
together with execution model details in Section~\ref{sec:disr_detaled_model}.
Finally, simulations are carried out with the opensource tool Nanoxim
to demostrate how DiSR preserves some of the main segment-based
properties and how compares to the tree-based approaches.

>>>>

% TODO: re-enable on long paper
%Exploring long term alternatives to the CMOS technology is gaining
%more and more revelance as the scaling trend of such devices keeps
%introducing new challenges. Power density, defect tolerance, testing
%costs and wire delays are only a few of the many critical aspects
%involved~\cite{itrs12}. While software parallelism and multicore
%approaches~\cite{horowitz2004, mudge2001, powell2009} are partially mitigating the impact of such
%constraints on performances, it is likely that the growing computing demand will
%eventually need even more radical architectural modifications and new
%paradigms in order to address the Computer Design challenges of the next
%decades.

In the last years, Self-assembled nanoscale architectures~\cite{winfree1998, yan2003}
are emerging as a promising technology due their tera/peta scale of
integration, defect tolerance tolerance and huge potential computing
capabilities. These technologies are certainly still at
their early stage of development, however different laboratory demos and
prof-of-concepts architectures have been presented~\cite{patwardhan2004, patwardhan2006, patwardhan2006_1, pistol2009}.
The main idea behind this approach is to exploit the physical regularity and
stabiliy of DNA structures in order to create a scaffold onto which
nano-devices (e.g. nanowires and nanotubes~\cite{bachtold2001, tans1998, cui2001}) can be
attached. This can be achieved by designing approapriate complementary DNA tags for
each terminal to be placed, so that a nano device will be attached
only where its own DNA tag matches a complementary tag on the DNA grid
scaffold (see figure~\ref{fig:dna_tag}.

\begin{figure}
  \centering
    \includegraphics[width=0.40\textwidth]{pictures/dna2b.eps}
  \caption{DNA tag matching}
  \label{fig:dna_tag}
\end{figure}

A detailed description of the chemical properties involved is far
beyond the scope of this paper (see also~\cite{braun1998, seeman1999}), so we will focus on
the three main challenges that this new fabrication process introduces
in Computer Design: (i)limited node complexity, (ii) large scale
randomness and (iii) high defect rates.

The \emph{limited node complexity} aspect is directly
related to the use of complementary DNA tags in order to place circuit
components. In a traditional CMOS process the complexity is introduced
using a photolithography mask, so that larger (more complex) circuits
only require larger masks; conversely, self-assembly achieves
complexity by increasing the number of unique DNA tags, since more
different tags means having more control on component placement. Ideally, by specifying a single
and unique tag for each nano transistor terminal, we could exactly
choose where each component would be placed. But the number of DNA
symbols forming the DNA sequence is limited (sequence of 4 nucleobases
G,A,T,C) and so creating many different tags (of a given lenght) would
mean make them more similar to each other, increasing the probability
of incorrect/partial matching. To avoid this problem we should limit
the number of unique tags, thus limiting the complexity at each node.
In particular, a budget of about $10,000$ transistor per node has been
estimated in~\cite{liu_jetcs}.

\emph{Large scale randomness} is the other fundamental condition of
self-assembled technology: DNA tags allow a control placement inside
node grid, but there is no control of where each of these  grids will
be placed on the whole network. As a consequence, other typical
properties of regular networks cannot be guaranteed, e.g. being
connected to a fixed number of neighbors, having a determined
orientation and so on.

Finally, \emph{defect rates}: while they are very hardly tolerated in mask
based top down design , the same nature of a bottom up self
assembled process cannot assume such a deterministic device placement
process.

These aspects of a DNA-self assembly process lead to some important
implications to be addressed when approaching to the design of
computer architecture: a computational model should be founded on a
distributed network of small computing and storage nodes, randomly
placed and interconnected (see figure~\ref{fig:nana}).
Proof-of-concept of such kind of computational model can be found for
example in~\cite{patwardhan2006_1}.  Since no regularity can be
assumed in such network, a topology agnostic strategy that avoids
deadlock should be adopted in order to route packets (e.g.
instructions and data operands) among nodes.  It is important to
underline that, altought we can assume an architecture similar to the
one depicted in~\ref{fig:nana} as background, the topology discovery
problem is more general and not stricly dependent on any of the
underlying computing architectures.

%where instructions and data operands travel
%on packets and routed using a topology-agnostic strategy that avoids
%deadlock, 
\begin{figure}
  \centering
    \includegraphics[width=0.20\textwidth]{pictures/dna1.eps}
  \caption{Irregular topology of computing and storage elements}
  \label{fig:nana}
\end{figure}

The contribution presented in this paper is to bring the advantages of
a segment-based approach to deadlock freedom without requiring any topology graph,
external defect map or centralized algorithm execution.  We introduce DiSR, a first effort for a Distributed
Segment-based Routing in large scale DNA self assembled networks.
While maintaining similar properties to segment-based , the DiSR
approach is not intended to discover the "optimal" segment choice
(ideally reachable with the knowledge of the topology graph) but just
to demonstrate a concrete model that can fit into such complex,
irregular and large sized topologies.
Given the irregular/high-defect rates scenario discussed above, is essential
avoiding non-scalable or resource hungry solutions such as
centralized off-chip algorithms, tree-based approaches, virtual channels
or expensive implementations based on hardware redundancy.

@@

The DiSR approach presented in this works is a first attempt of
reaching the same goal of SR (e.g. estabilishing segments) without
requiring a topology graph and a centralized exectution.

The paper is organized as follows: in Section~\ref{sec:disr_concepts}
we introduce the main elements of the DiSR approach, describing  
a draft of the DiSR model in Section~\ref{sec:disr_model}
together with execution model details in Section~\ref{sec:disr_detaled_model}.
In Section~\ref{sec:simulatio} simulations are carried out with the opensource tool Nanoxim
to demostrate how DiSR preserves some of the main segment-based
properties and how compares to the tree-based approaches. Finally a
draft of a VHDL hardware implementation is shown in
Section~\ref{sec:implementation} to evaluate the feasibility of the
approach in terms of node complexity.
